NodIO: a framework and architecture for pool-based evolutionary computation

### 2016-05-03 EvoSoft16/nodio-evosoft16.tex
Fixing some typos, closes #121

The sequence of events happens as follows:
\begin{enumerate}
\item First the {\sf NodIO} process is started in Node.js. As part of the
initialization procedure, the sequential number of experiments held in
a variable is initialized and a log
file is created. The shared pool implemented as hash that uses as keys
the chromosome strings is reset.
\item A volunteer follows the link to the experiment triggering a GET HTTP
request for static content and then the page is loaded.
\item The main script renders the page and creates the worker
  instance. In
order to create a dedicated worker instance, another GET request is sent to the
server, then the client receives the second worker implementation script.
Each worker instance executes over time one or more evolutionary algorithms
{\em islands}. Each island starts with a random population and it is assigned a
universally unique identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server so that that particular worker can be
identified and tracked in the logs. 
The initial parameters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following the previously
specified parameters. Once the population has been initialized, the EA
generation loop is started. After {\em n} generations  the chromosome
with the best fitness is sent to the shared 
population with a PUT HTTP request. If the best fitness is not the global best,
the EA loop must continue, but first a chromosome is randomly selected from
the shared pool. The current generation number and the best
chromosome are also sent to the client's main thread in order to update the DOM to
display the current state of the island (described next). These messages are
all asynchronous, so the EA loop of
the worker continues even before a message is received by the
target. If the best fitness is a global best, the best 
chromosome is sent to the server and the loop ends. A message is sent to the
main thread indicating that the best individual has been found.
\item The main thread receives messages from workers as events handled by
a callback function. When an iteration message is received, the page is
updated by plotting a dynamic graph of the generation number and best
fitness, and a representation of the chromosome is also generated.
\item When a global best is received from an island, the current experiment
ends, the experiment number is incremented, and the population array is reset.
\item When the main thread receives a message indicating that a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the start-up cost, only the parameters and population are reset and a new UUID
assigned.
\end{enumerate}
%

These two versions of the framework have been tested {\em in the
  wild} to measure its performance; results are published elsewhere,
but we also have to establish a baseline for comparison. This is only
for comparison purposes and not intended to prove that a volunteer
framework, if used exclusively, is faster than other
implementations. The main intention of this line of research is to
check how evolutionary algorithms work in this context, and also how
volunteer computing frameworks can be used to {\em supplement}
whatever resources, desktop computers, laptop or cloud, are availab.e

@@

These two experiments establish a baseline result and show that
the time to success depends on the population size, with a bigger population
contributing to have more diversity and thus speeding up the solution \cite{DBLP:conf/lion/LaredoDFGB13}. The volunteer computing
experiments that we will describe next do not and can not have the
same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic.

>>>>

The sequence of events happens as follows:
\begin{enumerate}
\item First the {\sf NodIO} process is started in Node.js. As part of the
initialization procedure, the sequential number of experiments held in
a variable is initialized and a log
file is created. The shared pool implemented as hash that uses as keys
the chromosome strings is reset.
\item A volunteer follows the link to the experiment triggering a GET HTTP
request for static content and then the page is loaded.
\item The main script renders the page and creates the worker
  instance. In
order to create a dedicated worker instance, another GET request is sent to the
server, then the client receives the second worker implementation script.
Each worker instance executes over time one or more evolutionary algorithms
{\em islands}. Each island starts with a random population and it is assigned a
universally unique identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server so that particular worker can be
identified and tracked in the logs. 
The initial parameters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following the previously
specified parameters. Once the population has been initialized, the EA
generation loop is started. After {\em n} generations  the chromosome
with the best fitness is sent to the shared 
population with a PUT HTTP request. If the best fitness is not the global best,
the EA loop must continue, but first a chromosome is randomly selected from
the shared pool. The current generation number and the best
chromosome are also sent to the client's main thread in order to update the DOM to
display the current state of the island (described next). These messages are
all asynchronous, so the EA loop of
the worker continues even before a message is received by the
target. If the best fitness is a global best, the best 
chromosome is sent to the server and the loop ends. A message is sent to the
main thread indicating that the best individual has been found.
\item The main thread receives messages from workers as events handled by
a callback function. When an iteration message is received, the page is
updated by plotting a dynamic graph of the generation number and best
fitness, and a representation of the chromosome is also generated.
\item When a global best is received from an island, the current experiment
ends, the experiment number is incremented, and the population array is reset.
\item When the main thread receives a message indicating that a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the start-up cost, only the parameters and population are reset and a new UUID
assigned.
\end{enumerate}
%

These two versions of the framework have been tested {\em in the
  wild} to measure its performance; results are published elsewhere,
but we also have to establish a baseline for comparison. This is only
for comparison purposes and not intended to prove that a volunteer
framework, if used exclusively, is faster than other
implementations. The main intention of this line of research is to
check how evolutionary algorithms work in this context, and also how
volunteer computing frameworks can be used to {\em supplement}
whatever resources, desktop computers, laptop or cloud, are available

@@

These two experiments establish a baseline result and show that
the time to success depends on the population size, with a bigger population
contributing to have more diversity and thus speeding up the solution searching \cite{DBLP:conf/lion/LaredoDFGB13}. The volunteer computing
experiments that we will describe next do not and can not have the
same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic.
